---
title: "Ex-Gauss for RT data"
output:
  html_document:
    df_print: paged
---

```{r message=FALSE}
library(tidyverse)
library(retimes)
library(boot)
library(broom)
library(lme4)
library(lmerTest)
library(sjPlot)

rtimes <- readRDS('rtimes_data.rds')
```


## Ex-Gaussian distribution parameters  
Estimating mu, sigma, tau for each participant and condition.  
mu: mean of Gaussian distribution  
sigma: SD of Gaussian distribution  
tau: tail (greater tau -> more long RTs -> positive skew)  

```{r}
sbj_params_ml <- rtimes %>%
  group_by(group, Subject, shape, congruency) %>% 
  summarise(timefit_out = list(data.frame(t(attr(retimes::timefit(TTime),"par"))))) %>%
                   unnest(cols = c(timefit_out)) %>% ungroup
```

### Parameter distributions
```{r fig.width=8, fig.height=8}
sbj_params_ml %>%
  gather(var, value, mu:tau) %>%
  ggplot(aes(value, fill=interaction(congruency, shape))) + 
  geom_density(alpha=.4) + 
  facet_wrap(var~group, ncol=2) + theme_bw()
```

## Fitting lmer with ex-Gaussian parameters 
```{r}
sbj_params_ml <- sbj_params_ml %>%
  mutate(group = relevel(factor(group), ref = "ctrl"),
         congruency = relevel(factor(congruency), ref = "Congruent"),
         shape = relevel(factor(shape), ref = "shape"))

options(contrasts = c("contr.sum","contr.poly"))
```
### mu
```{r}
mmu <- lmer(mu ~ group*shape*congruency + (1|shape:Subject) + (1|congruency:Subject), data=sbj_params_ml)
summary(mmu)
plot_model(mmu, type = "diag")[[4]]
```
### sigma
```{r}
msig <- lmer(sigma ~ group*shape*congruency + (1|shape:Subject) + (1|congruency:Subject), data=sbj_params_ml)
summary(msig)
plot_model(msig, type = "diag")[[4]]

```

### tau
```{r}
mtau <- lmer(tau ~ group*shape*congruency + (1|shape:Subject) + (1|congruency:Subject), data=sbj_params_ml)
summary(mtau)
plot_model(mtau, type = "diag")[[4]]

```

### tau (log model)
```{r}
mtau.log <- lmer(log(tau) ~ group*shape*congruency + (1|shape:Subject) + (1|congruency:Subject), data=sbj_params_ml)
summary(mtau.log)
plot_model(mtau.log, type = "diag")[[4]]
```


### Plot fixed effects
```{r}
new_data <- expand.grid(shape = c('shape', 'nonShape'), 
                      congruency = c('Congruent', 'InCon'), 
                      group = c('adhd', 'ctrl'))
                      
cbind(new_data, rtime = predict(mmu, newdata = new_data, re.form=NA)) %>%
  ggplot(aes(congruency, rtime, linetype=shape)) + geom_point() + 
  geom_line(aes(group=shape),size=1) +
  facet_wrap(~group) + theme_bw() + ggtitle("mu values")

cbind(new_data, rtime = predict(msig, newdata = new_data, re.form=NA)) %>%
  ggplot(aes(congruency, rtime, linetype=shape)) + geom_point() + 
  geom_line(aes(group=shape),size=1) +
  facet_wrap(~group) + theme_bw() + ggtitle("sigma values")

cbind(new_data, rtime = predict(mtau, newdata = new_data, re.form=NA)) %>%
  ggplot(aes(congruency, rtime, linetype=shape)) + geom_point() + 
  geom_line(aes(group=shape),size=1) +
  facet_wrap(~group) + theme_bw() + ggtitle("tau values")

cbind(new_data, rtime = predict(mtau.log, newdata = new_data, re.form=NA)) %>%
  ggplot(aes(congruency, rtime, linetype=shape)) + geom_point() + 
  geom_line(aes(group=shape),size=1) +
  facet_wrap(~group) + theme_bw() + ggtitle("tau values (log model)")

```


### Pairwise comparisons
Here for main effects only  
```{r}
difflsmeans(mmu, c("congruency", "shape", "group"))
difflsmeans(msig, c("congruency", "shape", "group"))
difflsmeans(mtau, c("congruency", "shape", "group"))
difflsmeans(mtau.log, c("congruency", "shape", "group"))

```

### Bootstrapping lmer estimates

```{r}
bootstrap.estimates <- function(model) {
  cb <- bootMer(model, FUN=fixef, nsim=1000 #.progress = "txt", PBargs = list(style=3)
                )
  cb
}
#cb$t %>% data.frame() %>%
#  gather(var, value) %>%
#  ggplot(aes(value)) + geom_density() + facet_wrap(~var, scales = "free_x")

bootstrap.estimate.plot <- function(cb, model) {
  boot_ci <- lapply(2:8, function(b) boot.ci(cb, type = "perc", index=b)$percent)

  ci <- do.call(rbind, boot_ci)
  td <- tidy(model)[2:8,]  
  td$conf.low <- ci[, 4]
  td$conf.high <- ci[, 5]

  g <- ggplot(td, aes(term, estimate)) +
    geom_point() +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width=.1) + coord_flip() +
    geom_hline(aes(yintercept=0), linetype= "longdash") + theme_bw()
  g

}
```


```{r}
 
cb.mmu <- bootstrap.estimates(mmu)
cb.msig <- bootstrap.estimates(msig)
cb.mtau <- bootstrap.estimates(mtau)
cb.mtau.log <- bootstrap.estimates(mtau.log)

bootstrap.estimate.plot(cb.mmu, mmu)
bootstrap.estimate.plot(cb.msig, msig)
bootstrap.estimate.plot(cb.mtau, mtau)
bootstrap.estimate.plot(cb.mtau.log, mtau.log)
```